Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-12-10 trunk/tex/rp.tex
spell check

%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{almdrsT} ends with
\RL{T} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherrent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

@@

{\bf Goals.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the fromal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will build a system using CATARM to check the strength
of the relations between internet documents viewed by a minor and
a set of Arabic text documents containing unrated or unwanted 
material. If the relation is strong, the system will flag access to
the document. 
All components and parts built during the development of solutions
to the case studies will be generalized and added to CATARM as 
library computational components. 
\end{itemize}

@@

The PI and his team at the American University of Beirut (AUB)
started work on an Arabic text mining framework~\cite{ATMine09}
that is available now online.
They have successfully developped ATSarf, a case-based morphological 
analyzer, and applied it on the hadith literature case study.

A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
We consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into
its content or \RL{matn} and its \noArRL{sanad} or the
chain of narrators.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names 
and once
connected toghether they form complex names of narrators. 
Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
name is \noTrRL{s`yd} as the word in between 
\noTrRL{bn} (son of) indicates.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyser.
We explain this approach in details in 
Section~\ref{s:design:method}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than two minutes.

@@

The Co-LPI, Rehab Duwairi, has several years of experience 
working on data mining for Arabic documents. 
She initially targeted Arabic text categorization where a 
dataset of 1000 documents where used to train and test 
classifiers~\cite{Duw06,Duw07}.
Word morphology comes to the surface when extracting the 
feature vectors of documents. 
In that work, words were stemmed before using them as features. 
Also, word frequency and inverted term frequency were used to 
quantify the importance of features.

In a recent work, Duwairi, Refae and Khasawneh~\cite{Duw09}
have addressed text categorization again. 
This research investigated possible word forms that can be 
used to create feature vectors of documents in particular: 
original word, stem of the word, light-stem (shallow stemming) 
of the word, and word augmented with a list of synonyms. 
Therefore, every document is represented using four different 
feature vectors, the result revealed that light-stemming is the 
best representation as it preserves the meaning of the words 
and yet reduces the dimensions of feature vectors. 
The dataset size was 15,000 documents that belong to three 
categories.

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:method}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologicaly equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

The biography books are fed to another MA unit controlled by 
a name and qualification pair controller that generates a 
map consisting of pairs of narrator names associated with 
qualifications. 
Then the graph of narrators and the narrator qualification map
are both processed through a locator component that computes
a graph coloring routine to color the graph of narrators. 
The color changes whener two consecutive narrators are far away
from each other in the text. 
The locator then looks at connected subgraphs of the same color,
and finds a median node in each of them to be the narrator 
associated with the qualifiers of the color. 
Then the locator also annotates the narrator with the 
corresponding qualifications. 
The result is an annotated partial order graph of narrators
with authentication qualification annotations.

@@

The diagram in Figure~\ref{f:security} describes our 
high level vision for what the solution of the 
security case study
will look like in the CATARM framework. 
The inputs are two separate sets of documents. 
The first set includes crime scene investigation reports, 
and the second set includes a 
interrogation reports with crime suspects as well as 
a database of previous interrogation reports. 
The CATARM system also takes as input a database of geographical
locations and a database of personal records. 
The aim is to check for inconsistencies in the iterrogation
reports and to identify possible connections between the suspects
and the crime scene.

@@

We feed $G_D, G_F, G_G,$ and $G_R$   
to a consistency checker.
The consistency checker considers the three dimensions 
$N$ names, $G$ locations, and $D$ dates and projects $G_R$
over each one of them.
For each dimension $k$, it computes a local measure 
$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
It flags the components in the graph where the accumelated 
measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
below a threshold $\theta$ as inconsistent.

@@

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morpholgical solution for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

The current use of morphological analyzers suffers from 
several problems. 
Analyzers that enumerate all possible partitions of a word
and match them against concatenative rules are computationaly 
expensive. 
Analyzers that exhaustively enumerate all possible solutions 
for a word may hurt performance and may
not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
For example, the first solution may be the one needed and 
we may not need to list the rest of the solutions. 
A white space delimited token may have more than one word
as is the case with ``run-on words''.

@@

Case-based analysis can guide the morphological analysis
to prune lots of the solutions and to bias the order of considered
decisions. 
For example, in the hadith literature case study pesented above, 
amongst other things, we are intersted in checking whether a 
word is a name.  
We can bias the order of considered decisions to consider the 
morphemes that can be names first. 
If one of them matches, we can ignore the rest of the solutions. 
In case a prefix is decided to be of a morphological category
that does not connect with a name, then we can just ignore
the analysis of the rest of the word. 
This is more efficient than generating all possible partitionings
of the word, checking for morphological correctness of each 
partition, and then inspecting the part of speech tags of each
correct solution. 
We desire to build a morphological analyzer that allows a 
case-based controller to intervene at every single decision
without much overhead.

\subsection{CATARM Query language}
\label{s:design:query}

@@

The state machine in Firgure~\ref{f:statemachine} controls 
the concatenation based state machines of the morphological
analyzer.
The transitions are excited
by inputs from the morphological analyzer. 
A NAME label represents a proper name, 
an NRC label represents a narrator connector such as
\RL{`an} "on behalf of", \RL{qaal} "said", 
or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} 
``the son of'' and represents a name connector.

@@

The diagram in Figure~\ref{f:chain} depicts our vision of 
an equivalent query.
This query generates the list of chains by first passing the 
hadith text into the morphological analyzer which 
extracts names and their position in text. 
We iterate over the extracted names and compute the distance.
Whenever the distance is less than a threshold $\theta_1$
we add the word to a temporary chain of names. 
Otherwise, we execute a sequential block to 
{\bf 1.} tabulate the chain into a temporary table containing 
the list of chains. 
{\bf 2.} Create a new chain whith the last element.
Finally, we filter those chains whose cardinality 
is less than some threshold  $\theta_2$.

@@

{\bf Simplify LCM based on the query.}
Given
L$L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to overapproximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

@@

We detailed an agressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years and projects a timeline
of work per quarter. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the status of the items 
and stand for 
{\bf P:} {\em planning phase}, 
{\bf W:} {\em work in progress}, 
{\bf D:} {\em deliverable}, 
{\bf M:}  {\em maintainance}, 
and 
{\bf E: } {\em enhance} respectively.

The $1,2,3,$ and $4$
superscripts of the deliverables in the 
in case studies row refer to the four 
case studies described in the propoal.

@@

In the following we list the budget items. 
Monatary values are provided in more details on the QNRF 
submission website.

\begin{itemize}
\item One research assistant at QU
\item Two research assistants at AUB, one of them will 
work partially from QU. 
\item Two month salary per year for the Co-LPI
\item Two month salary per year for the LPI
\item One course buy out per year for the Co-LPI
\item Two weeks consultant for the second and the third years
\item Licensing and membership fees to acquire digital libraries and corpora
\item High performance computer and storage devices
\item One conference for each PI per year
\item One conference for one RA from QU per year
\item One conference for one RA from AUB per year
\item Site visit to AUB 
\item Site visit to QU 
\item Other miscellanous items
\end{itemize}

@@

We will build a query language for relational Arabic text mining 
and a front-end to compile a user query into an optimized
computation. 
We will check how user friendly and how expressive the language is.
We will compare our query language with the sophisticated 
interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
We will look for other attempts to formalize Arabic text 
mining queries and we will also compare agains 
formal text mining queries from other text mining domains such
as the gene ontolgy domain~\cite{GeneOntology10},

@@

We anticipate CATARM to be able to successfully solve all the case
studies with resonable computational resources available to the 
average user. 
We will also generalize the components we will develop to solve
the case studies into a library of components that can be used
in other case studies.

@@

The completion of the CATARM framework will result in a 
significant contribution to the research community of 
NLP in general and Arabic Analysis in particular. 
The CATARM framework will act as vehicle to enhance and advance 
research in the area. 
We look forward for additional contributions to the field 
such as the following.
\begin{itemize}
\item We will explore developing and integrating vocalization
techniques. 
Vocalization techniques assign diacritics to unvocalized text. 
The prerequisites of a successful vocalizer are the accuracy of 
the morphological analyzer in addition to grammatical analyzers. 
\item We will explore the use of CATARM with more case studies
such as market analysis and opinion mining.
\item We will explore augmenting the lexicon of the
morphological analyzer with useful categorization and entity 
tags that reflect semantical meanings. 
These modifications would give CATARM additional capabilities in 
terms of allowing users to extract and use a greater variety of 
entities when building queries. 
The feedback of experts in the field of Arabic linguistics can 
continually add accuracy to the morphological solutions.  
\item We will explore further optimization techniques
that can help simplify and the analysis. 
\item We will explore generalizing our results to languages 
that resemble the Arabic language. 
Some semetic languages, such as the Amharic, cyrillic, and Hebrew,
share common features with the Arabic 
language. Some other languages are written using the 
Arabic script like Persian, Kurdish, Urdu, and Pashto. 
\item We will also investigate the case-based approach with
Latin languages and check how that compares with current techniques
in terms of accuracy and efficiency.
\item We will maintain CATARM and provide support to users and 
researchers of CATARM.
\end{itemize}

>>>>

%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{almdrsT} ends with
\RL{T} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

@@

{\bf Goals.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will build a system using CATARM to check the strength
of the relations between Internet documents viewed by a minor and
a set of Arabic text documents containing unrated or unwanted 
material. If the relation is strong, the system will flag access to
the document. 
All components and parts built during the development of solutions
to the case studies will be generalized and added to CATARM as 
library computational components. 
\end{itemize}

@@

The PI and his team at the American University of Beirut (AUB)
started work on an Arabic text mining framework~\cite{ATMine09}
that is available now online.
They have successfully developed ATSarf, a case-based morphological 
analyzer, and applied it on the hadith literature case study.

A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
We consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into
its content or \RL{matn} and its \noArRL{sanad} or the
chain of narrators.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names 
and once
connected together they form complex names of narrators. 
Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
name is \noTrRL{s`yd} as the word in between 
\noTrRL{bn} (son of) indicates.

@@

The framework now comprises a state of the art
case-based morphological analyzer, ATSarf, 
and computational components that allow the user to
specify his queries as one or more FSM
controlling the morphological analyzer.
We explain this approach in details in 
Section~\ref{s:design:method}.
ATSarf along with other computational components, 
extract the chains of narrators from a set of narrations
with above 95 percent accuracy and in less than two minutes.

@@

The Co-LPI, Rehab Duwairi, has several years of experience 
working on data mining for Arabic documents. 
She initially targeted Arabic text categorization where a 
data set of 1000 documents where used to train and test 
classifiers~\cite{Duw06,Duw07}.
Word morphology comes to the surface when extracting the 
feature vectors of documents. 
In that work, words were stemmed before using them as features. 
Also, word frequency and inverted term frequency were used to 
quantify the importance of features.

In a recent work, Duwairi, Refae and Khasawneh~\cite{Duw09}
have addressed text categorization again. 
This research investigated possible word forms that can be 
used to create feature vectors of documents in particular: 
original word, stem of the word, light-stem (shallow stemming) 
of the word, and word augmented with a list of synonyms. 
Therefore, every document is represented using four different 
feature vectors, the result revealed that light-stemming is the 
best representation as it preserves the meaning of the words 
and yet reduces the dimensions of feature vectors. 
The data set size was 15,000 documents that belong to three 
categories.

@@

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:method}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologically equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

The biography books are fed to another MA unit controlled by 
a name and qualification pair controller that generates a 
map consisting of pairs of narrator names associated with 
qualifications. 
Then the graph of narrators and the narrator qualification map
are both processed through a locator component that computes
a graph coloring routine to color the graph of narrators. 
The color changes when two consecutive narrators are far away
from each other in the text. 
The locator then looks at connected subgraphs of the same color,
and finds a median node in each of them to be the narrator 
associated with the qualifiers of the color. 
Then the locator also annotates the narrator with the 
corresponding qualifications. 
The result is an annotated partial order graph of narrators
with authentication qualification annotations.

@@

The diagram in Figure~\ref{f:security} describes our 
high level vision for what the solution of the 
security case study
will look like in the CATARM framework. 
The inputs are two separate sets of documents. 
The first set includes crime scene investigation reports, 
and the second set includes a 
interrogation reports with crime suspects as well as 
a database of previous interrogation reports. 
The CATARM system also takes as input a database of geographical
locations and a database of personal records. 
The aim is to check for inconsistencies in the interrogation
reports and to identify possible connections between the suspects
and the crime scene.

@@

We feed $G_D, G_F, G_G,$ and $G_R$   
to a consistency checker.
The consistency checker considers the three dimensions 
$N$ names, $G$ locations, and $D$ dates and projects $G_R$
over each one of them.
For each dimension $k$, it computes a local measure 
$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
It flags the components in the graph where the accumulated 
measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
below a threshold $\theta$ as inconsistent.

@@

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morphological solution for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

The current use of morphological analyzers suffers from 
several problems. 
Analyzers that enumerate all possible partitions of a word
and match them against concatenative rules are computationally 
expensive. 
Analyzers that exhaustively enumerate all possible solutions 
for a word may hurt performance and may
not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
For example, the first solution may be the one needed and 
we may not need to list the rest of the solutions. 
A white space delimited token may have more than one word
as is the case with ``run-on words''.

@@

Case-based analysis can guide the morphological analysis
to prune lots of the solutions and to bias the order of considered
decisions. 
For example, in the hadith literature case study presented above, 
amongst other things, we are interested in checking whether a 
word is a name.  
We can bias the order of considered decisions to consider the 
morphemes that can be names first. 
If one of them matches, we can ignore the rest of the solutions. 
In case a prefix is decided to be of a morphological category
that does not connect with a name, then we can just ignore
the analysis of the rest of the word. 
This is more efficient than generating all possible partitions
of the word, checking for morphological correctness of each 
partition, and then inspecting the part of speech tags of each
correct solution. 
We desire to build a morphological analyzer that allows a 
case-based controller to intervene at every single decision
without much overhead.

\subsection{CATARM query language}
\label{s:design:query}

@@

The state machine in Figure~\ref{f:statemachine} controls 
the concatenation based state machines of the morphological
analyzer.
The transitions are excited
by inputs from the morphological analyzer. 
A NAME label represents a proper name, 
an NRC label represents a narrator connector such as
\RL{`an} "on behalf of", \RL{qaal} "said", 
or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} 
``the son of'' and represents a name connector.

@@

The diagram in Figure~\ref{f:chain} depicts our vision of 
an equivalent query.
This query generates the list of chains by first passing the 
hadith text into the morphological analyzer which 
extracts names and their position in text. 
We iterate over the extracted names and compute the distance.
Whenever the distance is less than a threshold $\theta_1$
we add the word to a temporary chain of names. 
Otherwise, we execute a sequential block to 
{\bf 1.} tabulate the chain into a temporary table containing 
the list of chains. 
{\bf 2.} Create a new chain with the last element.
Finally, we filter those chains whose cardinality 
is less than some threshold  $\theta_2$.

@@

{\bf Simplify LCM based on the query.}
Given
L$L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to over-approximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

@@

We detailed an aggressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years and projects a timeline
of work per quarter. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the status of the items 
and stand for 
{\bf P:} {\em planning phase}, 
{\bf W:} {\em work in progress}, 
{\bf D:} {\em deliverable}, 
{\bf M:}  {\em maintenance}, 
and 
{\bf E: } {\em enhance} respectively.

The $1,2,3,$ and $4$
superscripts of the deliverables in the 
in case studies row refer to the four 
case studies described in the proposal.

@@

In the following we list the budget items. 
Monetary values are provided in more details on the QNRF 
submission website.

\begin{itemize}
\item One research assistant at QU
\item Two research assistants at AUB, one of them will 
work partially from QU. 
\item Two month salary per year for the Co-LPI
\item Two month salary per year for the LPI
\item One course buy out per year for the Co-LPI
\item Two weeks consultant for the second and the third years
\item Licensing and membership fees to acquire digital libraries and corpora
\item High performance computer and storage devices
\item One conference for each PI per year
\item One conference for one RA from QU per year
\item One conference for one RA from AUB per year
\item Site visit to AUB 
\item Site visit to QU 
\item Other miscellaneous items
\end{itemize}

@@

We will build a query language for relational Arabic text mining 
and a front-end to compile a user query into an optimized
computation. 
We will check how user friendly and how expressive the language is.
We will compare our query language with the sophisticated 
interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
We will look for other attempts to formalize Arabic text 
mining queries and we will also compare against 
formal text mining queries from other text mining domains such
as the gene ontology domain~\cite{GeneOntology10},

@@

We anticipate CATARM to be able to successfully solve all the case
studies with reasonable computational resources available to the 
average user. 
We will also generalize the components we will develop to solve
the case studies into a library of components that can be used
in other case studies.

@@

The completion of the CATARM framework will result in a 
significant contribution to the research community of 
NLP in general and Arabic Analysis in particular. 
The CATARM framework will act as vehicle to enhance and advance 
research in the area. 
We look forward for additional contributions to the field 
such as the following.
\begin{itemize}
\item We will explore developing and integrating vocalization
techniques. 
Vocalization techniques assign diacritics to unvocalized text. 
The prerequisites of a successful vocalizer are the accuracy of 
the morphological analyzer in addition to grammatical analyzers. 
\item We will explore the use of CATARM with more case studies
such as market analysis and opinion mining.
\item We will explore augmenting the lexicon of the
morphological analyzer with useful categorization and entity 
tags that reflect semantical meanings. 
These modifications would give CATARM additional capabilities in 
terms of allowing users to extract and use a greater variety of 
entities when building queries. 
The feedback of experts in the field of Arabic linguistics can 
continually add accuracy to the morphological solutions.  
\item We will explore further optimization techniques
that can help simplify and the analysis. 
\item We will explore generalizing our results to languages 
that resemble the Arabic language. 
Some Semitic languages, such as the Amharic, Cyrillic, and Hebrew,
share common features with the Arabic 
language. Some other languages are written using the 
Arabic script like Persian, Kurdish, Urdu, and Pashto. 
\item We will also investigate the case-based approach with
Latin languages and check how that compares with current techniques
in terms of accuracy and efficiency.
\item We will maintain CATARM and provide support to users and 
researchers of CATARM.
\end{itemize}
